Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-01 article.Rmd
improving intro about german situation

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover these fees, sometimes also referred to as article-processing charges (APC), authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. Yet, how and to what extent these research support activities are effective in terms of the number of supported articles and associated costs remains under debate.

One reason why studying how much was spent on publication fees is in most cases difficult is that payments for these charges are fragmented across the budgets of funding agencies, research institutions, and libraries, or, if support is limited, are taken from personal budgets. Another key problem for the empirical study of institutional spending on publication fees is that funding for open access publications lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to a complex landscape of variable pricing schemes [@Pinfield_2015] include submission or page charges [@BJ_RK_2012]. Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

Previous empirical studies asking how much was being spent for publishing in open access journals confirmed that authors make use of funding or institutional support for covering open access publication fees. In 2010, the SOAP project carried out a comprehensive survey and asked 9,645 authors from various disciplines how they financed publication fees [@soap]. Of the respondents, 12 % had used personal budgets. By contrast, most had access to institutional support structures or research funding. These findings are consistent with that of other surveys, adding that sources of funding exists mostly in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Asked about the amount spent on publication fees, 22,977  respondents in the SOAP survey mostly reported prices between 501 and 1000 € (12,9%), and between 1001 and 3000 € (9.9 %). However, more than half of the authors answering this question had not to pay a publication fee when publishing in open access journals, and 14 % did not report the amount spent. While examining the indexing coverage of fully open access journals in subject-specific databases, @Walters_2011 collected publication fees as journal characteristic in 2009. This study reported an average fee of 923 $. Journals indexed in the Web of Science charged 1,171 $ on average. @Solomon_2012 calculated an similar average of 906 $ for publishing in fully open access journals. However, they found a large price variation between 8 to 3,900 $. They suggested to cluster fully open access journals using publication fees into several groups. In this regard, while one set consists of costly high-impact journals (2000 $ - 4000 $), a large number of journals charge 1500 $ - 2000 $. Journals with broad scope like PLOS ONE form another segment charging between 1000 - 1500 $, just as mid-priced journals from commercial publishers (500 - 1000 $), society journals (below 500 $), and low-priced journals (below 200 $).

Complementary to survey data or the consultation of journal web pages for determining spending on publication fees, some research funders and research performing institutions have recently begun to disclose their expenditures on publication fees as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. The latter was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variations, as well as variations in the number of articles the analysed UK universities supported between 2007 - 2014.

<!-- This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Some European initiatives have therefore begun to collect and disclose their expenditures on open access journal articles as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report expenditures along with bibliographic information, including title, journal and publisher, persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).-->

@@

This paper focuses on how much German universities and research organization spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has been encouraging open access publishing since years. The DFG launched its "Open-Access Publishing" program in 2009. This programme has strongly influenced how university libraries support open access publication fees through funds.[^3] To reduce administrative burdens, grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers. Grantees are also required to report the institutional publication output and their publication fee spending to the DFG on a regular basis, and to present the university-wide strategy to sustain the funds when DFG’s initial support runs out lately in 2019.

The DFG enforces a set of criteria grantees have to comply with, leading to similar implementations for supporting open access publishing across German universities: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding VAT) [@Fournier_2013]. Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this funding program, contributing to the diversity of schemes in Germany. In response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing. The Allianz der Wissenschaftsorganisationen[^10c], representing all major research organizations in Germany, marks in this regard transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payments in May 2014. After putting its approach to the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] as a basis for discussion, the library invited others to participate under the umbrella of the Open APC initiative, and contributions from other German research institutions followed soon after.

@@

Drawing on self-reported data from this open data initiative, the aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. The analysis focused on the amount of institutional payments for publication fees, and compared these findings with those from related Austrian and UK initiatives. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

>>>>

The rise of open access journals matches the increasing relevance of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. To cover publication fees, also referred to as article-processing charges (APC), authors tend to make use of funding that grant agencies or academic institutions provide [@Suber_2012]. Yet, how and to what extent these research support activities are effective in terms of the number of supported articles and associated costs remains under debate.

One reason why studying how much was spent on publication fees is in most cases difficult is that payments for these charges are fragmented across the budgets of funding agencies, research institutions, and libraries, or, if support is limited, are taken from personal budgets. Another key problem is that funding for open access journals using publication fees lacks transparency because the parties involved - authors, universities, funders, publishers - neither release information on who pays for what nor the costs of publishing [@Bj_rk_learned_2014], a situation similar to the lack of transparency regarding journal subscriptions [@Lawson_Meghreblian_2015]. It also remains unclear which factors contribute to price formation.[^4b] While fixed prices for individual articles are common, agreements between publishers and institutions often provide discounts, and publishers sometimes waive publication fees for authors from low-income countries [@BJ_RK_2012; @Lawson_waiver_2015]. Other factors leading to variable pricing schemes include submission or page charges [@BJ_RK_2012]. Hybrid journals substantially add to this complexity of open access funding [@Kingsley_2014]. These journals, allowing articles to be published immediately as open access after a charge was paid, rely both on subscriptions and publication fees as revenue sources. Although the uptake of open access through hybrid journals was described as lower and more expensive compared to that of fully open access journals [@Solomon_2012; @Bj_rk_learned_2014], this model has gained attention through recent science policies, notably because of open access policies from the UK [@Pinfield_Review_2015].

Previous empirical studies asking how much was being spent for publishing in open access journals confirmed that authors make use of funding or institutional support for covering publication fees. In 2010, the SOAP project carried out a comprehensive survey in which 9,645 authors from various disciplines provided insights into how they financed publication fees [@soap]. The majority of the respondents used institutional support structures or research funding. By contrast, 12 % paid publication fees individually. These findings are consistent with that of other surveys, adding that sources of funding exists mostly in higher income countries, mainly to support research articles in the bio- and physical sciences [@Solomon_2011]. Personal budgets, on the other hand, are likely used to cover low price publication fees [@Solomon_2011; @Bj_rk_2015].

Asked about the amount being spent on publication fees, 22,977 SOAP survey respondents reported prices between 501 and 1000 € (12,9%), and between 1,001 and 3,000 € (9.9 %). However, more than half of the authors answering this question had not to pay a publication fee to get publish in an open access journal, and 14 % did not report the amount spent. While examining the indexing coverage of fully open access journals in bibliographic databases in 2009, @Walters_2011  also provided publication fees as journal characteristics. Their study reported an average publication fee of 923 $, and for Journals indexed in the Web of Science 1,171 $ on average. @Solomon_2012 calculated an similar average of 906 $ for publishing in fully open access journals. However, they found a large price variance across journals and articles. Accordingly, they suggested to cluster fully open access journals using publication fees into several groups. In this regard, while one set consisted of costly high-impact journals (2,000 $ - 4,000 $), a large number of journals charged 1,500 $ - 2,000 $. Journals with broad scope like PLOS ONE formed another segment charging between 1,000 - 1,500 $, just as mid-priced journals from commercial publishers (500 - 1,000 $), society journals (below 500 $), and low-priced journals (below 200 $).

Complementary to survey data or the consultation of journal web pages for determining spending on publication fees, some research funders and research performing institutions have recently begun to disclose their spending on publication fees as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report payments made along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. Parts of Jisc cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as variations in the number of articles UK universities supported between 2007 - 2014.

<!-- This complex situation of fee-based open access publishing creates difficulties for researchers and practitioners alike. Some European initiatives have therefore begun to collect and disclose their expenditures on open access journal articles as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc] and the Austrian Science Fund FWF [@fwf_apc]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report payments made along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version, and a link to a deposit in a subject repository. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates. In the case of the Wellcome Trust, crowd-sourcing data cleaning activities through a Google spreadsheet in combination with checks against bibliographic sources massively improved the spending data (see comments in @wellcome_apc).-->

@@

This paper focuses on how much German universities and research organization spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced through its "Open-Access Publishing" program how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. The DFG has enforced a set of criteria grantees have to comply with, leading to similar implementations for supporting open access publishing across German universities: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax). Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers. [@Fournier_2013].

Research institutes organized in the Fraunhofer-Gesellschaft, Helmholtz-Gemeinschaft, Leibniz-Gemeinschaft, and Max-Planck-Gesellschaft are not eligible for this funding program, contributing to the diversity of schemes in Germany. But in response, some organizations have adopted similar processes to support authors. The Max-Planck-Gesellschaft operates their long-lasting open access activities, including handling spending and publisher agreements centrally, through the Max Planck Digital Library [@Schimmer_2013; @Sikora_2015], while the Leibniz-Gemeinschaft set up a dedicated open access fund in 2016.

The evolving institutional support structures to cover open access publication fees has led to calls for an unified approach towards supporting open access journal publishing. The Allianz der Wissenschaftsorganisationen[^10c], a science policy board representing all major research organizations in Germany, marks cost transparency as one way to sustain an "adequate open access publication system" [@allianz]. Reflecting Austrian and UK initiatives to share institutional spending on open access publication fees as open data, as well as professional discussions on open access publishing, Bielefeld University Library began to openly share its payments in May 2014. After engaging with the working group "Electronic Publishing" of the Deutsche Initiative für Netzwerkinformation (DINI)[^11] other German institutions joined under the umbrella of the Open APC initiative soon after.

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these findings with those from related Austrian and UK initiatives. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.
